Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-15 ieee_paper.tex
Updated figures in text. See end of text.
formatting
text of section 4

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. (We will show that an implementation relying on
compiler vectorization does not lead to desirable performance.) The
registers in the MIC architecture are 512 bits wide and can store 16
floats or integers. (In comparison to OpenCL or CUDA, one can think of
these registers as a warp.) The end goal is to be able to load and
format the non zero values $a,b,c,d$ and the vector entries $p,q,r,s$
in two vectorial register to be able to apply Fused Multiply-Add on
them, which is depicted in Figure~\ref{fig:tensor_product}. Bringing
the data in the vector register as depicted in the figure is the
difficult part that we now explain. A thread will perform the
multiplications one row at a time, and parallelism is achieved by
giving blocks of rows to each core usign an OpenMP construct.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes (sometimes called channels) of 128 bits
which are made of 4 elements of 32 bits. A swizzling operation allows
to reorder the elements within each lane. While a permutation allow to
change the orders of the lanes. Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast.

@@

Loading the entries from the {\tt x} vector is a little bit more
complicated.

We use the vector operations presented in
Figure~\ref{fig:MIC-vect-instr} in our code, which we'll explain in
the text.

\def\loadps{\ttt{\_mm512\_load\_ps}}
\def\loadds{\ttt{\_mm512\_load\_ds}}
\def\fmadps{\ttt{\_mm512\_fmad\_ps}}
\def\gatherps{\ttt{\_mm512\_i32gather\_ps}}
\def\swizzleps{\ttt{\_mm512\_swizzle\_ps}}
\def\storenrngops{\ttt{\_mm512\_storenrngo\_ps}}
\def\castsi{\ttt{\_mm512\_castsi512\_ps}}
\def\permute{\ttt{\_mm512\_permute4f128\_epi32}}
\def\intmask{\ttt{\_mm512\_int2mask}}
\def\loadunpack{\ttt{\_mm512\_mask\_loadunpacklo\_epi32}}
\def\castitops{\ttt{\_mm512\_castsi512\_ps}}
\def\castpstoi{\ttt{\_mm512\_castps\_si512\_ps}}

\begin{figure*}
  %
  \begin{center}
    \begin{tabular}{|l|l|}
      \hline
      \loadps &  load 16 consecutive floats to register\\
      \fmadps &  multiply/add of 16 floats\\
      \gatherps &  gather 16, possibly disconnected floats from memory\\
      \swizzleps &  reorder a channel, and duplicate across channels\\
      \storenrngops &  store 16 floats to memory without register rewrite or reordering\\
      \castpstoi & reinterpret 16 floats as integers\\
      \castitops & reinterpret 16 integers as floats\\
      \permute &  permute channels; do not change individual channels\\
      \intmask &  \\
      \loadunpack &  \\
      \hline
    \end{tabular}
  \end{center}
  %
  \caption{List of MIC vectorial instructions used.}
\label{fig:MIC-vect-instr}
\end{figure*}
In the vector instructions, \ttt{ps} refers to a float (4 bytes),
while \ttt{epi32} refers to a 4-byte integer. Although we do not
discuss double precision in this paper, \ttt{ds} refers to an 8 byte
double precision real number (i.e., \loadds).  Each vector register is
broken up into four channels of 128 bits each. It is possible to
interchange these channels at a cost of a "few" \qes{exact numbers?}
cycles, and it is also possible to execute swizzle operations within a
channel. For example, if the 128 bytes of each channel are labelled as
$ABCD$, the vector instruction \ttt{\_mm512\_swizzle\_ps(v,
  \_MM\_SWIZ\_REG\_AAAA)} replaces each channel by $AAAA$.

Figure~\ref{code:read_aaaa} is a function that reads in four floats
(a,b,c,d) and creates the 16-float vector dddd,cccc,bbbb,aaaa:

@@

    //three more times with BBBB, CCCC, and DDDD
  }
  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
} 
\end{verbatim}
\caption{mat mul code}
\label{code:mat_mul}
\end{figure*}

Note that the ``permute'' function is using the {\tt
  \_mm512\_permute4f128\_epi32} instruction and therefore typecast are
necessary to convert to a vector of int type.

@@

>>>>

We discuss how the implementation of the v4m4 kernel works using
vectorial instruction. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show that an implementation
relying on compiler vectorization does not lead to desirable
performance.) The registers in the MIC architecture are 512 bits wide
and can store 16 floats or integers. (In comparison to OpenCL or CUDA,
one can think of these registers as a warp.) The end goal is to be
able to load and format the non zero values $a,b,c,d$ and the vector
entries $p,q,r,s$ in two vectorial register to be able to apply Fused
Multiply-Add on them, which is depicted in
Figure~\ref{fig:tensor_product}. Bringing the data in the vector
register as depicted in the figure is the difficult part that we now
explain. A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
usign an OpenMP construct.

@@

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of 4 lanes (sometimes called channels) of 128 bits
which are made of 4 elements of 32 bits. A swizzling operation allows
to reorder the elements within each lane. While a permutation allow to
change the orders of the lanes. Figure~\ref{fig:swizzling_permutation}
illustrates these two possibilities. Both permutation and swizzling
support common reordering and broadcast. In the code, the ``permute''
function needs to use the {\tt \_mm512\_permute4f128\_epi32}
instruction which is only defined on integers and therefore typecast
are necessary to convert to a vector to/from the integer type.

@@

Loading the entries from the {\tt x} vector is a little bit more
complicated. First the columns pointers of 4 different non zeroes are
loaded in a vector register and the values are distributed one per
lane of the vector register using an {\tt unpack}
operation\footnote{Notice that there are two calls to unpack for the
  LSB and MSB. Both are mandatory even according the documentation of
  the hardware}. They replicated so that each fill a vector lane using
a swizzle operation. Then each value is multiplied by 4 (because there
are 4 vectors) and the offsets $(3,2,1,0)$ is added each lane to
obtain the correct index of the elements of the {\tt x} vector within
each lane. Then a {\tt gather} operation is performed bring the 16
entries of the {\tt x} vector into the SIMD register. At this point we
have a SIMD register where each lane is the four vector entries for a
non zero element. Using a broadcast permutation, one can replicate an
entry of the four vectors accross each lane.

@@

    //three more times with BBBB, CCCC, and DDDD
  }
  _mm512_storenrngo_ps(dom.result_vt+nb_mat*nb_vec*r, accu);  
} 
\end{verbatim}
\caption{mat mul code.}
\label{code:mat_mul}
\end{figure*}

@@

\input{additional_figures}
